Effective Reionization Likelihood from Planck 2018 Data (?)

### 2021-03-04 paper/rei.tex

...

@@

%
%
\subsection{Reionization Principal Components}
\label{sec:PC}
We begin by parametrizing $x_e(z)$, the ionization fraction relative to the fully ionized hydrogen at redshift $z$, into its principal components $S_{a}(z)$ with respect to the CMB $E$-mode polarization:
%
\begin{equation}
x_e(z)=\xef(z)+\sum_{a}m_{a}S_{a}(z),
\label{eq:mmutoxe}
\end{equation}
%
where $m_a$ are the PC amplitudes and $\xef(z)$ is the fiducial model. We obtain the PCs $S_{a}(z)$ as eigenfunctions of the Fisher information matrix for $x_e(z)$ in a given range $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements 
%
\beq
F_{ij} = \sum_l \frac{1}{\sigma_l^2} \frac{\partial C_l^{EE}}{\partial x_e(z_i)}\frac{\partial C_l^{EE}}{\partial x_e(z_j)} = \sum_a S_a(z_i) \sigma_a^{-2} S_a(z_j),
\eeq
%
where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the variances of the PCs. We have rank-ordered the PCs from low to high variance, so that for the range $z_{\rm min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.~\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit. In the work that follows, we therefore truncate the sum at $a = 1 .. n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ for the fiducial analysis. 
We also test its robustness by taking 
 for $z_{\rm max} = 50$ where $n_{\rm{PC}} = 7$ is required at the cosmic variance limit.
[Maybe insert reference to a plot of all five PCs (we might refer to its shapes later)]
Note that the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
\begin{equation}
m_{a}=
  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin},
\label{eq:xetommu}
\end{equation}
%
where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision. We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
For the Planck data set, most of the information on the ionization history comes from the first two modes which probe the amount of low vs.\ high redshift optical depth.   However all 5 PCs for $z_{\rm max}=30$ carry some constraint, with the higher modes controlling finer variations in redshift.

We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}, where the PCs were discretized at $\delta z = 0.25$, and computed at the fiducial model $x_e^{\rm fid} = 0.15$. Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the $\zmax = 30$ PCs to be computed with Planck 2015 rather than WMAP best-fit parameters, which results in minor differences between the PC functions used in Ref.~\cite{Heinrich:2016ojb} and here. For $z<6$, we follow CAMB and assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a width $\Delta z = 0.5$ (see also \S \ref{sec:example1}).

@@

 \begin{figure}
          \includegraphics[width=0.95\columnwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
          \caption{Reionization PCs for $\zmax = 30$.} 
          \label{fig:xe}
\end{figure}

@@

\subsection{Effective PC Likelihood}
\label{sec:KDE}
The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood that can be used to test any reionization model with the CMB without a cumbersome reanalysis at the level of power spectra data. In the following, we briefly recap the kernel density estimate technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$. Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~(\ref{eq:xetommu}). Since $\mathbf{m}$ can take any continuous value, we approximate its effective likelihood with a kernel density estimate
\beq
{\cal L}_{\rm PC}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
\eeq
where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$. To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx 1.1 \times 10^{6}$ chain samples. 
% 1.06 cut w/ physicality prior
% 1.072 no cut
For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient.
%Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors. 
Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
\begin{equation}
P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p).
\end{equation}

@@

Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
\begin{equation}
\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
\label{eq:cumtau}
\end{equation}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.  The underlying reason is that CMB observables probe the optical depth and whether it is weighted toward high or low redshift through the amplitude and shape of 
$C_\ell^{EE}$ -- the higher the redshift, the
larger the relative contribution at higher $\ell$.

Specific models essentially place an implicit prior on the redshift dependence of the cumulative optical depth.  
%For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
% \begin{equation}
%x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
% \label{eqn:tanh}
% \end{equation}
% where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
% excludes finite ionization above its transition redshift by definition.
The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no range-bound prior constraints on the mode amplitudes $m_a$ are imposed. The one subtlety when placing constraints on the cumulative optical depth is that the analysis also allows
unphysical ionization fractions where $0 \le x_e \le x_e^{\rm max}$  is not satisfied.
Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
 we do impose a physicality prior
 by truncating the posteriors of $m_a$
 {\it after} obtaining them from Markov Chain Monte Carlo sampling, following Ref.~\cite{Mortonson:2008rx}
%
\begin{equation}
\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
\label{eq:prior_sum}
\end{equation}
where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
\begin{equation}
m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
\label{eq:individualprior}
\end{equation}
\ch{It turns out that in practice the prior on the sum of squares in Eq.~\ref{eq:prior_sum} is weaker than the individual priors in Eq.~\ref{eq:individualprior} for the Planck 2018 data.}
\wh{comment on the sum prior?  we don't seem to talk about it later - is it applied and is it always weaker in practice?}
\ch{CH: The sum prior is applied at all times; I also checked  (for the zmax = 30 chains this year) that it is weaker than the prior on individual m's just as in Planck 2015.}
%
Note that the original $0\le x_e \le x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality.  
In other words, no physical model is excluded by these priors but some unphysical models are included.  In \S \ref{sec:note_on_priors} we further explore the role of these and other choices of PC priors, and in \S\ref{sec:effective_likelihood} we test PC constraints against direct constraints for two example models.

@@

\caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$. We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}. The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle. The physicality priors on the ionization history used in this work are delineated by the box boundaries. [TODO: update legend] [Add text: The black solid line indicates the steplike models varying $\tau$. make point that tanh ML is now within the the center of PC constraints][say that Planck 2015 point is chain ML, whereas Plnack 2018 point is from minimizer, also is a different tanh model (w/ dz = 0.015(1+z), not dz = 0.5)]}
\label{fig:plot_mjs_2018_vs_2015}
\end{figure*}
%results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf
%
In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$. The shaded region in the $m_1-m_2$ plane represents the parameter region that would be excluded by an additional 
physicality prior from Eq.~(\ref{eq:individualprior}), whereas for the other cases the box itself represents this prior. Results using the Planck 2018 data are shown in blue. While all five PC amplitudes are  constrained by Planck, the first two are particularly well-constrained. Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth.  \wh{This preference also increases the impact of placing a physicality prior at low values of $m_1$.
We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}.
In this excluded region, the optical depth of models are dominated by the low redshift end, but due to the fixed ionization at $z>z_{\rm min}=6$, the desired total optical depth cannot be achieved without an unphysical ionization elsewhere.} \ch{[Not sure if we should say ``In this excluded region" referring to the entire gray band, or specify overlap with data constraints where $m_2 < 0$ mostly?]}

In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show
the Planck 2018 constraints using the older \texttt{simall\_EE} likelihood.  The improvement 
from switching to the \texttt{srollv2} likelihood mainly strengthens the upper bound in the best constrained direction in the $m_1-m_2$ plane which shifts constraints toward higher \ch{[lower?]} total optical depth. \ch{[Or maybe you meant Planck 2015 to simall EE?]}

@@

\begin{figure}
\includegraphics[width=0.40\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $\zmax=30$ PC plane, the $m_1-m_2$ plane, comparing Planck 2015 (gray) vs Planck 2018 results using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) vs the \texttt{srollv2} likelihood (blue), the latter had better foreground removal and is the one we show as our fiducial results throughout this paper. 
\wh{some of these figure versions look like low res bit maps - final versions should be vector graphics and look uniform in style between figures.} \ch{got it, will polish for sure.}
}
\label{fig:plot_m1m2_2015_vs_2018}
\end{figure}

@@

Using Eqs.~(\ref{eq:mmutoxe}) and (\ref{eq:cumtau}), we can derive constraints on the total optical
depth $\tau=\tau(0,z_{\rm max)$ with the physicality priors of Eq.~\ref{eq:individualprior}.  These are shown in Fig.~XX \wh{figure could include $\tau$, $\tau_{12}$ and possibly $\tau_{12}$ without physicality priors?} and correspond to
\beq
\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068},
\eeq  
\wh{comment on comparison to tanh tau, flexknot etc.}
Notice that the PC mean, while still compatible with the tanh 
tau derived constraints, is almost $0.5\sigma$ higher...

@@

In Fig.~\ref{fig:plot_taugtz_2015_vs_2018} top panel, we compare the 68\% and 95\% confidence level contours in $\tau(z, \zmax)$ for the Planck 2015 (red regions) and the Planck 2018 results (solid lines) using all 5 PCs.
\wh{remind the reader about impact of PC filtering smoothing out features in redshift - refer to the tanh examples to illustrate}
In the lower panel, we compare the Planck 2018 results for the two different low-$\ell$ $EE$ likelihoods.   Notice first the dramatic improvement in the
constraint at high-$z$.  Whereas the Planck 2015 allowed and even preferred
finite optical depth at $z\gtrsim 15$, Planck 2018 no longer does....
On the other hand, some optical depth is still allowed 
\beq
\tau_{\rm PC}(15, 30) < 0.023 \; (95\%\; \mathrm{C.L.})\;\;(\texttt{srollv2}),
\eeq
in contrast with the tanh model where it is strictly forbidden in the direct analysis and in the PC filtered effective analysis only enters due to the smoothness of the first few PCs in redshift space.
\wh{contrast this with Flexknot and talk about compatibility with example 2.}

% with this prior you still have that $\tau(>z)$ plot includes unphysical models (however, most likely total tau doesn't change, and if any bias from unphysical models, it would lower high-z tau).\\

We also show using a separate PC chains with $\zmax = 50$ that there is no evidence for optical depth at $z>30$. 
In Fig.~\ref{fig:plot_taugtz_zmax30_vs_zmax50}, we show that 68\% and 95\% C.L. contours of $\tau(z, \zmax)$ are consistent between the $\zmax = 30$ (red regions) and $\zmax = 50$ results (black lines).

>>>>

We publish Relike, an fast and accurate effective likelihood code based on constraints from the latest Planck 2018 data on the CMB reionization principal components (PC). This code allows one to constrain arbitrary global ionization history $x_e(z)$ between $6 < z < 30$ with Planck without having to modify CAMB each time a different model is tested. We tested the code on two example models
%: 1) the tanh model with a single transition redshift, and 2) a two-parameter model with two transitions allowing for a high-redshift ionization plateau. 
which showed good agreement with directly 
%The Relike chains converged 300$\times$ and 100$\times$ faster for model 1 and 2 respectively, and showed good agreement compared to directly
sampling the original Planck likelihoods.
We also find that contrary to the Planck 2015, the Planck 2018 PC posteriors are well approximated by a Gaussian in PC amplitude space, this allows us to replace the kernel density estimate (KDE) evaluation of the PC chains in the code with an even faster Gaussian evaluation. We supply both the KDE and Gausisan modes of the likelihood in a CosmoMC implementation, and the Gaussian mode alone in an easy-to-use standalone python likelihood package that can be connected to any other sampler. We expect the code release to dramatically reduce the time for testing individual ionization models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as 21cm, luminosity function, star formation history, etc.
%We expect the code release to be useful for individual model testing as the user would not need to modify CAMB every time a different ionization history model is needed; the effective likelihood code also makes it easier to perform joint constraints of the Planck data with other reionization data such as 21cm, luminosity functions, star formation rate measurements, etc.
Finally, we derive model-independent constraints: $\tau_{\rm PC} = ...$ and $\tau_{\rm PC}(15, 30) < XX $ (95\% C.L.). The upper limits on high-redshift optical depth is at least a factor of 2 looser those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method; we demonstrate robustness of our results by explicitly recovering  $\tau_{\rm PC}(15, 30) < XX $ (95\% C.L.) using a  two-parameter toy model without using PCs.

@@

%
%
 \begin{figure}
          \includegraphics[width=0.95\columnwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
          \caption{Reionization PCs for $\zmax = 30$.} 
          \label{fig:xe}
\end{figure}

\subsection{Reionization Principal Components}
\label{sec:PC}
We begin by parametrizing $x_e(z)$, the ionization fraction relative to the fully ionized hydrogen at redshift $z$, into its principal components $S_{a}(z)$ with respect to the CMB $E$-mode polarization:
%
\begin{equation}
x_e(z)=\xef(z)+\sum_{a}m_{a}S_{a}(z),
\label{eq:mmutoxe}
\end{equation}
%
where $m_a$ are the PC amplitudes and $\xef(z)$ is the fiducial model. We obtain the PC basis functions $S_{a}(z)$ as eigenfunctions of the Fisher information matrix for $x_e(z)$ in a given range $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements 
%
\beq
F_{ij} = \sum_l \wh{\frac{2 l+1}{2} \frac{\partial \ln C_l^{EE}}{\partial x_e(z_i)}\frac{\partial \ln C_l^{EE}}{\partial x_e(z_j)}}= \sum_a \frac{ S_a(z_i) S_a(z_j)}{\sigma_a^2},
\eeq
%
where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the variances of the PCs.
The Fisher matrix is computed at a fiducial model which we take to be $x_e^{\rm fid}(z)=0.15$ in the parameterized range.
 Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the PCs to be computed with Planck 2015 rather than WMAP best-fit 5 remaining $\Lambda$CDM parameters, which results in minor differences between the PC functions used in Ref.~\cite{Heinrich:2016ojb} and here.

We then rank-ordered the PCs from low to high variance, so that for the range $z_{\rm min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.~\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit. In the work that follows, we therefore truncate the sum at $a = 1 .. n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ for the fiducial analysis.\footnote{We also test the robustness of our fiducial analysis in \S XX by  taking 
 for $z_{\rm max} = 50$ where $n_{\rm{PC}} = 7$ is required at the cosmic variance limit.}

 In Fig.~\ref{fig:xe} (top), we show the corresponding fiducial basis functions $S_a(z)$.  
Notice that the series resembles a Fourier decomposition of $x_e(z)$ rank ordered from low to high frequency.
This is because high frequency variations in $x_e$ leave little observable imprint in $C_l^{EE}$.  
As a consequence  the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
\begin{equation}
m_{a}=
  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin},
\label{eq:xetommu}
\end{equation}
%
where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision. We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
For the Planck data set, most of the information on the ionization history comes from the first two modes which probe the amount of low vs.\ high redshift optical depth.   However all 5 PCs for $z_{\rm max}=30$ carry some constraint, with the higher modes controlling finer variations in redshift.

We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}. 
For $z<6$, we follow CAMB and assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a transition width $\Delta z = 0.5$ (see also \S \ref{sec:example1}).

@@

@@

\subsection{Effective PC Likelihood}
\label{sec:KDE}
The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood that can be used to test any reionization model with the CMB without a cumbersome reanalysis at the level of power spectra data. In the following, we briefly recap the kernel density estimate (KDE) technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$. Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~(\ref{eq:xetommu}). Since $\mathbf{m}$ can take any continuous value, we approximate its effective likelihood with a kernel density estimate
\beq
{\cal L}_{\rm PC}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
\eeq
where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$. To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx 1.1 \times 10^{6}$ chain samples. 
% 1.06 cut w/ physicality prior
% 1.072 no cut
For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient.
%Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors. 
Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
\begin{equation}
P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p).
\end{equation}

@@

Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
\begin{equation}
\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
\label{eq:cumtau}
\end{equation}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.  
\wh{Possibly add a few words on the PC version and refer to the
bottom panel of Fig.~\ref{fig:xe}}
The underlying reason is that CMB observables probe the optical depth and whether it is weighted toward high or low redshift through the amplitude and shape of 
$C_\ell^{EE}$ -- the higher the redshift, the
larger the relative contribution at higher $\ell$.

Specific models essentially place an implicit prior on the redshift dependence of the cumulative optical depth.  
%For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
% \begin{equation}
%x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
% \label{eqn:tanh}
% \end{equation}
% where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
% excludes finite ionization above its transition redshift by definition.
The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no range-bound prior constraints on the mode amplitudes $m_a$ are imposed. The one subtlety when placing constraints on the cumulative optical depth is that the analysis also allows
unphysical ionization fractions where $0 \le x_e \le x_e^{\rm max}$  is not satisfied.
Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
 we do impose a  prior
 by truncating the posteriors of $m_a$
 {\it after} obtaining them from Markov Chain Monte Carlo sampling, following Ref.~\cite{Mortonson:2008rx}
%
\begin{equation}
\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
\label{eq:prior_sum}
\end{equation}
where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
\begin{equation}
m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
\label{eq:individualprior}
\end{equation}
In practice the prior on the sum of squares in Eq.~(\ref{eq:prior_sum}) is strictly weaker than the individual priors in Eq.~(\ref{eq:individualprior}) for
% I think it can't depend on the data but does depend on the truncation
%the Planck 2018 data \wh{and
our fiducial analysis.  Henceforth we refer to Eq.~(\ref{eq:individualprior}) as the physicality prior.}

%
Note that the original $0\le x_e \le x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for a physical ionization history.  
In other words, no physical model is excluded by these priors but some unphysical models are included.  In \S \ref{sec:note_on_priors} we further explore the role of these and other choices of PC priors, and in \S\ref{sec:effective_likelihood} we test PC constraints against direct constraints for two example models.

@@

\caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$. We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}. The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle. The physicality priors on the ionization history used in this work are delineated by the box boundaries \wh{and the unshaded region for $m_1$}. [TODO: update legend] [Add text: The black solid line indicates the steplike models varying $\tau$. make point that tanh ML is now within the the center of PC constraints][say that Planck 2015 point is chain ML, whereas Plnack 2018 point is from minimizer, also is a different tanh model (w/ dz = 0.015(1+z), not dz = 0.5)] \wh{double check relative sign on the definition of $m_i$ between 2015-18 - seems like some correlations have flipped, e.g. $m_4$.} }
\label{fig:plot_mjs_2018_vs_2015}
\end{figure*}
%results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf
%
In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$. The shaded region in the $m_1-m_2$ plane represents the parameter region that would be excluded by an additional 
physicality prior from Eq.~(\ref{eq:individualprior}), whereas for the other cases the box itself represents this prior. Results using the Planck 2018 data are shown in blue. While all five PC amplitudes are  constrained by Planck, the first two are particularly well-constrained. Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth.  \wh{This preference also increases the impact of placing a physicality prior at low values of $m_1$.
We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}.
In this excluded region, the optical depth of models are dominated by the low redshift end, but due to the fixed ionization at $z>z_{\rm min}=6$, the desired total optical depth cannot be achieved without an unphysical ionization elsewhere.} \ch{[Not sure if we should say ``In this excluded region" referring to the entire gray band, or specify overlap with data constraints where $m_2 < 0$ mostly?]}

In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show
the Planck 2018 constraints using the older \texttt{simall\_EE} likelihood.  The improvement 
from switching to the \texttt{srollv2} likelihood mainly strengthens the upper bound in the best constrained direction in the $m_1-m_2$ plane which shifts constraints toward higher total optical depth.

@@

\begin{figure}
\includegraphics[width=0.40\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $\zmax=30$ PC plane, the $m_1-m_2$ plane, comparing Planck 2015 (gray) vs Planck 2018 results using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) vs the \texttt{srollv2} likelihood (blue), the latter had better foreground removal and is the one we show as our fiducial results throughout this paper. 
}
\label{fig:plot_m1m2_2015_vs_2018}
\end{figure}

@@

Using Eqs.~(\ref{eq:mmutoxe}) and (\ref{eq:cumtau}), we can derive constraints on the total optical
depth $\tau=\tau(0,z_{\rm max})$ with the physicality priors of Eq.~\ref{eq:individualprior}.  These are shown in Fig.~XX \wh{figure could include $\tau$, $\tau_{12}$ and possibly $\tau_{12}$ without physicality priors?} and correspond to
\beq
\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068} \;\; (\texttt{srollv2}),
\label{eq:tau_pc_zmax30}
\eeq  
\wh{comment on comparison to tanh tau, flexknot etc.}
Notice that the PC mean, while still compatible with the tanh 
tau derived constraints, is almost $0.5\sigma$ higher...

@@

\ch{[CH: Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior} needs to go somewhere]}
% with this prior you still have that $\tau(>z)$ plot includes unphysical models (however, most likely total tau doesn't change, and if any bias from unphysical models, it would lower high-z tau).\\

In Fig.~\ref{fig:plot_taugtz_2015_vs_2018_simallEE_vs_2018_srollv2} top panel, we compare the 68\% and 95\% confidence level contours in $\tau(z, \zmax)$ for the Planck 2015 (red regions) and the Planck 2018 results (solid lines) using all 5 PCs.
\wh{remind the reader about impact of PC filtering smoothing out features in redshift - refer to the tanh examples to illustrate}
In the lower panel, we compare the Planck 2018 results for the two different low-$\ell$ $EE$ likelihoods.   Notice first the dramatic improvement in the
constraint at high-$z$.  Whereas the Planck 2015 allowed and even preferred
finite optical depth at $z\gtrsim 15$, Planck 2018 no longer does....
On the other hand, some optical depth is still allowed 
\beq
\tau_{\rm PC}(15, 30) < 0.023 \; (95\%\; \mathrm{C.L.})\;\;(\texttt{srollv2}),
\eeq
in contrast with the tanh model where it is strictly forbidden in the direct analysis and in the PC filtered effective analysis only enters due to the smoothness of the first few PCs in redshift space.
\wh{contrast this with Flexknot and talk about compatibility with example 2.}

We also show using a separate PC chains with $\zmax = 50$ that there is no evidence for optical depth at $z>30$. 
In Fig.~\ref{fig:plot_taugtz_zmax30_vs_zmax50}, we show that the 68\% and 95\% C.L. contours of $\tau(z, \zmax)$ are consistent between the $\zmax = 30$ (red regions) and $\zmax = 50$ results (black lines).
